---
title: "Final Exam"
author: "Ainsley Carlson"
date: "2024-12-16"
output: word_document
---




## Question 1
```{r}
#set working directory and read in data
setwd("C:/Users/OwNeR/OneDrive - West Chester University of PA/Desktop/Stat MS/Fall 2024/Stat 503")
final<-read.csv("FinalData.csv")
```
### Part A
```{r}
str(final)
```

There are 100 observations of three variables in this dataset. The variable names are "x", "y", and "var.1" with "x" and "var.1" being numeric while "y" is a character.

### Part B
```{r}
final[4,1]+final[12,3]
```
The value of the fourth row and first column and the twelfth row and third column is `r final[4,1]+final[12,3]`. 


### Part C
```{r}
head(final[,2],10)
```
### Part D
```{r}
mean(final[,3], na.rm=TRUE)
```

The mean of the non-missing values in the third column is `r `mean(final[,3], na.rm=TRUE)`.


### Part E
```{r}
prop.table(table(final$y))
```


## Question 2
### Part A
```{r}
final$var1.na<-is.na(final[,3])
```

```{r}
sum=0
new.vec<-rep(0,nrow(final))

for (i in 1:nrow(final)) {
  if (final[i,4]==TRUE) {
    new.vec[i]=0
  } else if (final[i,4]==FALSE & final[i,3]<0) {
    new.vec[i]=final[i,3]*(-1)
  } else {
    new.vec[i]=final[i,3]+2
  }
  sum=sum+new.vec[i]
}

```


### Part B
```{r}
new.vec <- ifelse(is.na(final$var.1), 0, 
                  ifelse(final$var.1 <0, 
                  final[,3]*(-1), 
                  final[,3]+2))

sum(new.vec)              
```




## Question 3
### Part A
```{r, message=FALSE}
#load tidyverse
library(tidyverse)
```

### Part B
```{r}
final %>%
  mutate(dif=var.1-x) %>%
  filter(y!="A") %>%
  group_by(y)  %>%
  summarize(mean=mean(dif, na.rm=TRUE)) %>%
  arrange(desc(mean))

```

